%=========================================================================
% EVALUATION
%=========================================================================
\section{Evaluation}
\label{sec:eval}
In this section, we will analyze the protection benefits and overheads introduced by our methodology.
%
We have implemented the hardware components of our design by making extensions to the AVR instruction set architecture.
%
The VHDL model of the extended processor is synthesizable.
%
We have instantiated the processor on Xilinx Vertex 2 Pro XC2VP30 FPGA.
%
Our performance overheads are measured using Modelsim 6.0 simulator.
%
The software library and applications were compiled using \texttt{avr-gcc} cross compiler.
%
%\subsection{Micro-Benchmarks}
%
\subsection{Performance Overhead}
%
We first present micro-benchmarks that measure CPU overhead introduced by the protection mechanism.
%
Table~\ref{tab:microbmperf} contains the overhead of run-time checks present in our mechanism.
%
We compare our overhead with a completely software based approach to memory protection through binary rewrites proposed in~\cite{ram07harbor} for the AVR architecture.
%
The software based approach also introduces identical run-time checks except that they are implemented in assembly language without any modifications to the processor architecture.
%
The results clearly indicate the superior performance of run-time checkers implemented in hardware.
%
\begin{table}[htdp]
\centering
\small{
\begin{tabular}{|l|c|c|}
	\hline
	Function Name & AVR Extension & AVR Binary Rewrite\\
	\hline
	Memmap Checker & 1 & 65\\
	Cross Domain Call & 5 & 65\\
	Cross Domain Ret & 5 & 28\\
	Save Ret Addr & 0 & 38\\
	Restore Ret Addr & 0 & 38\\
	\hline
\end{tabular}}
\caption{Overhead (CPU cycles) of Memory Protection Routines}
\label{tab:microbmperf}
\end{table}
%

The high overhead of software based memory map checker is mainly due to complex bit shift operations that are required to translate write addresses to memory map lookup.
%
Cross domain call and return have an overhead of five clock cycles when implemented in hardware.
%
The overhead occurs because the current domain identity, stack bound and return address have to be pushed to the safe stack before they can be updated with new values.
%
The total information that needs to be pushed to the stack is five bytes and only one byte can be written every clock cycle.
%
Similarly on the cross domain return, the five clock cycles are expended in restoring the values read from the safe stack.
%
Saving and restoring return addresses to the safe stack does not introduce any added overhead.
%
This is because the hardware unit for safe stack simply takes over the address bus when the processor is pushing the return address to the run-time stack.
%
By stealing the address bus from the processor, the hardware unit is able to simply redirect the store of the return addresses to the safe stack.

Next we evaluate the software library.
%
Performance overhead is also introduced by updates to memory map during allocation, free and transfer of memory within the system.
%
Table~\ref{tab:malloc_comparison} compares the overhead of memory allocation routines in the presence and absence of the protection mechanism.
%
Relatively higher overhead of \texttt{change\_own} and \texttt{free} calls is due additional checks that are introduced to prevent illegal ownership transfer or freeing of memory blocks by non-owners.
%
\begin{table}[htdp]
\centering
\small{
\begin{tabular}{|l|c|c|}
	\hline
	Function Name & Normal & Protected \\
	\hline
	malloc  & 343 & 610\\
	free & 138 & 425\\
	change\_own & 55 & 365 \\
	\hline
\end{tabular}}
\caption{Overhead (CPU cycles) of memory allocation routines}
\label{tab:malloc_comparison}
\end{table}
%

\subsection{Resource Utilization}
%
Resource utilization can be partitioned into sections.
%
The resource utilization of the software library and the overhead of hardware checkers.
%

%
%Code memory usage increases by about 15\% in protected kernel relative to unprotected kernel.
%
%Increase is mainly due to memory map API and jump table used by cross domain call mechanism.
%
%There is no significant change in program memory usage going from two protection domains to multiple protection domains.
%
%Data memory usage increases by 5\% and 9.5\% in two-domain and multi-domain systems respectively, relative to unprotected kernel.
%
%This is the maximum possible overhead as the memory map stores layout and ownership information of entire address space.
%
Code and data memory usage of the software library is shown in Table~\ref{tab:swlibsize}.
%
Maximum memory map size is 256 bytes for multi-domain protection.
%
This represents an overhead of 6.25\%.
%
However, by modifying data layout, portion of address space that requires memory map for protection can be reduced.
%
For example, memory map can be configured only to protect the heap and safe-stack.
%
By abutting these two data-structures, size of memory map required can be reduced to 140 bytes for multi-domain protection.
%
For two domain protection, the overhead can be reduced to only 70 bytes (1.7\%).
%
The total code memory usage of the software library is only 3674 bytes (2.8\%).
%
\begin{table}[htdp]
\centering
\small{
\begin{tabular}{|l|c|c|}
	\hline
	SW Component & FLASH (B) & RAM (B)\\
	\hline
	Dynamic Memory & 1204 & 2054\\
	Memory Map & 422 & 256 \\
	Jump Table & 2048 & 0 \\
	\hline
\end{tabular}}
\caption{FLASH and RAM overhead of software library}
\label{tab:swlibsize}
\end{table}
%

The hardware overhead of our mechanism is shown in Table~\ref{tab:hwsize}.
%
These results were computed by synthesizing our processor on Xilinx ISE 8.2i.
%
Most of the additions to the core area are in the memory map decoder that maintains a barrel shifter to support arbitrary bit-shifts in a single clock cycles.
%
We can eliminate this overhead if the processor is synthesized for a fixed block size and number of protection domains.
%
The overall increase in the core area is about 32\%.
%
This represents a modest increase in the overall area of the chip as the core occupies only a small fraction of the overall area.
%
Bulk of the chip area is occupied by SRAM and FLASH memories.
%
\begin{table}[htdp]
\centering
\small{
\begin{tabular}{|l|c|c|}
	\hline
	HW Component & Ext. Gate Count & Orig. Gate Count\\
	\hline
	AVR Core & 22498 & 16419\\
	Fetch Decoder & 6783 & 6685\\
	MMC & 2284 & N/A \\
	Safe Stack & 1749 & N/A \\
	Domain Tracker & 541 & N/A \\
	\hline
\end{tabular}}
\caption{Gate count overhead of hardware extensions}
\label{tab:hwsize}
\end{table}
%






